---
title: "Page 3"
output: 
  html_document:
    toc: true
    toc_float: true
---

Devon's Workspace
# Header
text here 

## Subsection 1
text/code here

## Subsection 2
text/code here



```{r}
 library(haven)
library(tidyverse)
library(scales)
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)
library(plotly)
library(kableExtra)
```


```{r}
liver_data_raw = read_sas("/Users/devonpark/Library/CloudStorage/OneDrive-ColumbiaUniversityIrvingMedicalCenter/tx_li.sas7bdat")

```

GENERAL:

```{r}
liver_data_1 = liver_data_raw |> 
  mutate(tx_date = as.Date(REC_TX_DT),
        listing_date = as.Date(CAN_LISTING_DT),
        tx_year = year(tx_date),
        days_listed = as.numeric(difftime(tx_date, listing_date, units = "days")),
        alcohol_subtype = case_when(
                CAN_DGN %in% c(4215, 4216) | CAN_DGN2 %in% c(4215, 4216) ~ "Alcohol-related Cirrhosis",
                CAN_DGN == 4217 | CAN_DGN2 == 4217 ~ "Acute Alcohol-related Hepatitis",
                TRUE ~ "Other")) %>% # days on waitlist
filter(!is.na(PERS_ID) & !is.na(tx_date) & tx_year >= 2002 & REC_AGE_AT_TX >= 18,
      alcohol_subtype != "Other") |> # MELD era only, Adults only
arrange(PERS_ID, tx_date) |> 
  janitor::clean_names()



liver_data_clean_1 = 
  liver_data_1 |>
    mutate(
      don_race_srtr_std = don_race_srtr,
      don_race_srtr_std = case_when(
        don_race_srtr_std == "ASIAN"   ~ "Asian",
        don_race_srtr_std == "WHITE"   ~ "White",
        don_race_srtr_std == "BLACK"   ~ "Black or African American",
        don_race_srtr_std == "NATIVE"  ~ "American Indian or Alaska Native",
        don_race_srtr_std == "MULTI"   ~ "Multi-Racial",
        don_race_srtr_std == "PACIFIC" ~ "Native Hawaiian or Other Pacific Islander",
        don_race_srtr_std == ""        ~ "NOT SPECIFIED",
        TRUE ~ don_race_srtr_std #For all rows that did not match any of the previous recoding conditions, keep the original value
      )
    )
```


```{r}
liver_data_clean_1 |>
  count(tx_year, don_race_srtr_std) |>
  group_by(tx_year) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = tx_year, y = pct, fill = don_race_srtr_std)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Donor Race by Transplant Year",
    x = "Transplant Year",
    y = "Percent",
    fill = "Donor Race"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-------------------------------------------------------------------

```{r}
liver_data = liver_data_raw |> 
  mutate(tx_date = as.Date(REC_TX_DT),
        listing_date = as.Date(CAN_LISTING_DT),
        tx_year = year(tx_date),
        days_listed = as.numeric(difftime(tx_date, listing_date, units = "days")),
        alcohol_subtype = case_when(
                CAN_DGN %in% c(4215, 4216) | CAN_DGN2 %in% c(4215, 4216) ~ "Alcohol-related Cirrhosis",
                CAN_DGN == 4217 | CAN_DGN2 == 4217 ~ "Acute Alcohol-related Hepatitis",
                TRUE ~ "Other")) %>% # days on waitlist
filter(!is.na(PERS_ID) & !is.na(tx_date) & tx_year >= 2020 & REC_AGE_AT_TX >= 18,
      alcohol_subtype != "Other") |> # MELD era only, Adults only
arrange(PERS_ID, tx_date) |> 
  janitor::clean_names()
```

Looking at race:
```{r}
#Clean don_Race column

liver_data_clean = 
liver_data |> 
  mutate(
    
    don_race_raw = don_race,#Clean original race codes
    true_missing = is.na(don_race_raw) | don_race_raw %in% c("NA", "", " "), # 2. Identify TRUE missing before numeric conversion
   
    don_race_num = suppressWarnings(as.numeric(trimws(as.character(don_race_raw)))),  #Converts to numeric (will produce NA for non-numeric multi-racial codes)

    don_race = case_when(
      don_race_num == 8 ~ "White",
      don_race_num == 16 ~ "Black or African American",
      don_race_num == 32 ~ "American Indian or Alaska Native",
      don_race_num == 64 ~ "Asian",
      don_race_num == 128 ~ "Native Hawaiian or Other Pacific Islander",
      don_race_num == 256 ~ "Arab or Middle Eastern",
      don_race_num == 512 ~ "Indian Sub-continent",
      don_race_num == 1024 ~ "Unknown (for Donor Referral only)",
      don_race_num == 2000 ~ "Hispanic/Latino",

      don_race_num %in% c(24, 40, 48, 56, 72, 80, 88, 96, 104, 112, 120, 
                          136, 144, 192, 200, 232) ~ "Multi-Racial",

      true_missing ~ "Missing", #see if there are TRUE missing BEFORE numeric conversion

      is.na(don_race_num) ~ "Multi-Racial", #Numeric conversion failures → Multi-Racial (your new rule)

      TRUE ~ as.character(don_race_raw) #Fallback in case
    )
  ) 
  
```


```{r}
#liver_data_clean |> 
#  distinct(don_race_srtr)

liver_data_clean = 
  liver_data |>
    mutate(
      don_ty_factor = case_when(
      don_ty == "C" ~ "Deceased Donor",
      don_ty == "L" ~ "Living Donor",
      TRUE          ~ "Other / Unknown"),
    don_ty_factor = factor(don_ty_factor, levels = c("Deceased Donor", "Living Donor", "Other / Unknown")),
    
      don_race_srtr_std = don_race_srtr,
      don_race_srtr_std = case_when(
        don_race_srtr_std == "ASIAN"   ~ "Asian",
        don_race_srtr_std == "WHITE"   ~ "White",
        don_race_srtr_std == "BLACK"   ~ "Black or African American",
        don_race_srtr_std == "NATIVE"  ~ "American Indian or Alaska Native",
        don_race_srtr_std == "MULTI"   ~ "Multi-Racial",
        don_race_srtr_std == "PACIFIC" ~ "Native Hawaiian or Other Pacific Islander",
        don_race_srtr_std == ""        ~ "NOT SPECIFIED",
        TRUE ~ don_race_srtr_std #For all rows that did not match any of the previous recoding conditions, keep the original value
      )
    )
  

```

Race breakdown since 2020 annually for all donors 
```{r}
liver_data_clean |>
  count(tx_year, don_race_srtr_std) |>
  group_by(tx_year) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = tx_year, y = pct, fill = don_race_srtr_std)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Donor Race by Transplant Year",
    x = "Transplant Year",
    y = "Percent",
    fill = "Donor Race"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


Side by side Deceased vs Living Donor Race Break Down (combining years 2020-2025)
```{r}
interactive_race = liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>
  count(don_ty_factor, don_race_srtr_std) |>
  group_by(don_ty_factor) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = don_ty_factor, 
             y = pct, 
             fill = don_race_srtr_std,
             text = paste0(
      "Donor Type: ", don_ty_factor, "<br>",
      "Race: ", don_race_srtr_std, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n
    ))) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Race Composition by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Race"
  )+
  theme_minimal()
```

Interactive Version:
```{r}
ggplotly(interactive_race, tooltip = "text")
```


Exploratory: Look at deceased vs living donors and gender

```{r}
liver_data_clean |> count(don_ty)
liver_data_clean |> count(don_gender)
liver_data_clean |> filter(don_ty == "L" | don_ty == "C") |> count(don_gender)
liver_data_clean |> filter(don_ty == "C") |> count(don_gender)
liver_data_clean |> filter(don_ty == "L") |> count(don_gender)


liver_data_clean |> count(don_relationship_ty)
```

Exploratory: Look at citizenship
```{r}
liver_data_clean |> count(don_citizenship)
#916 NA values

#Code	Description
#1	1: US Citizen
#2	2: Resident Alien
#3	3: Non-Resident Alien, Specify Country
#4	4: Non-US Citizen/US Resident
#5	5: Non-US Citizen/Non-US Resident
#998	998: Unknown
```

```{r}

liver_data_clean <- liver_data_clean |>
  mutate(
    don_citizenship_label = case_when(
      don_citizenship == 1 ~ "US Citizen",
      don_citizenship == 2 ~ "Resident Alien",
      don_citizenship == 3 ~ "Non-Resident Alien (Specify)",
      don_citizenship == 4 ~ "Non-US Citizen/US Resident",
      don_citizenship == 5 ~ "Non-US Citizen/Non-US Resident",
      don_citizenship == 998 ~ "Unknown",
      is.na(don_citizenship) ~ "Missing",
      TRUE ~ "Other"
    ),
    don_ty_label = case_when(
      don_ty == "C" ~ "Deceased Donor",
      don_ty == "L" ~ "Living Donor",
      TRUE          ~ "Other / Unknown"
    )
  )
```

Stacked barchart of citizenship: Decesased v Living Donor
```{r}

liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>   # only deceased + living
  count(don_ty_label, don_citizenship_label) |>
  group_by(don_ty_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = don_ty_label, y = pct, fill = don_citizenship_label)) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Citizenship Status by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Citizenship"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11),
    legend.position = "right"
  )
```

Interactive plotly version
```{r}

plotly_citizenship = liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>   # only deceased + living
  count(don_ty_label, don_citizenship_label) |>
  group_by(don_ty_label) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(
    x = don_ty_label,
    y = pct,
    fill = don_citizenship_label,
    text = paste0(
      "Donor type: ", don_ty_label, "<br>",
      "Citizenship: ", don_citizenship_label, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n
    )
  )) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Citizenship Status by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Citizenship"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11),
    legend.position = "right"
  )


ggplotly(plotly_citizenship, tooltip = "text")
```

Interactive: Donor Gender
```{r}
gender_interactive = liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>
  count(don_ty_factor, don_gender) |>
  group_by(don_ty_factor) |>
  mutate(pct = n / sum(n),
        don_gender = case_when(
        don_gender == "M" ~ "Male",
        don_gender == "F" ~ "Female",
        TRUE ~ don_gender
    )) |>
  ungroup() |>
  ggplot(aes(
    x = don_ty_factor,
    y = pct,
    fill = don_gender,
    text = paste0(
      "Donor Type: ", don_ty_factor, "<br>",
      "Gender: ", don_gender, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n
    )
  )) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor Gender Composition by Donor Type",
    x = "Donor Type",
    y = "Percent of Donors",
    fill = "Gender"
  ) +
  theme_minimal()

ggplotly(gender_interactive, tooltip = "text")
```
------------------------------

RECIPIENT INFO


CLEAN THE DATA
```{r}
liver_data_clean = liver_data_clean |>
  mutate(
    can_gender = factor(
      can_gender,
      levels = c("F", "M"),
      labels = c("Female", "Male")),
    can_race_srtr_std = can_race_srtr,
    can_race_srtr_std = case_when(
      can_race_srtr_std == "ASIAN"   ~ "Asian",
      can_race_srtr_std == "WHITE"   ~ "White",
      can_race_srtr_std == "BLACK"   ~ "Black or African American",
      can_race_srtr_std == "NATIVE"  ~ "American Indian or Alaska Native",
      can_race_srtr_std == "MULTI"   ~ "Multi-Racial",
      can_race_srtr_std == "PACIFIC" ~ "Native Hawaiian or Other Pacific Islander",
      can_race_srtr_std == ""        ~ "NOT SPECIFIED",
      TRUE                           ~ can_race_srtr_std
    )
  )

```

Race dsitribution from 2020 to 2025
```{r}
race_donor_int = liver_data_clean |>
  count(tx_year, can_race_srtr_std) |>
  group_by(tx_year) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = tx_year, 
             y = pct, 
             fill = can_race_srtr_std,
             text = paste0(
      "Year: ", tx_year, "<br>",
      "Race: ", can_race_srtr_std, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n)
      )) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of Recipient Race by Transplant Year",
    x = "Transplant Year",
    y = "Percent of Recipients",
    fill = "Race"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

ggplotly(race_donor_int, tooltip = "text")
```

Gender of Recipients
```{r}
gender_donor_int = liver_data_clean |>
  count(can_gender) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(x = "All Recipients", 
             y = pct, 
             fill = can_gender,
             text = paste0(
      "Gender: ", can_gender, "<br>",
      "Percent: ", scales::percent(pct, accuracy = 0.1), "<br>",
      "Count: ", n)
      )) +
  geom_col(color = "white") +
  scale_y_continuous(labels = percent_format()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Overall Recipient Gender Composition",
    x = "",
    y = "Percent of Recipients",
    fill = "Gender"
  ) +
  theme_minimal()

ggplotly(gender_donor_int, tooltip = "text")
```

Age distribution of Donors and Recipients 

```{r}
liver_data_clean |>
  select(don_age, rec_age_at_tx) |>
  tidyr::pivot_longer(cols = everything(),
                      names_to = "type",
                      values_to = "age") |>
  mutate(
    type = factor(type,
                  levels = c("don_age", "rec_age_at_tx"),
                  labels = c("Donor", "Recipient"))
  ) |>
  ggplot(aes(x = age, fill = type)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Comparison of Donor and Recipient Age Distributions",
    x = "Age (years)",
    y = "Density",
    fill = "Group"
  ) +
  theme_minimal()
```


Age difference between donor and recipients
```{r}
liver_data_clean = liver_data_clean |>
  mutate(
    age_diff = don_age - rec_age_at_tx   # positive means donor older than recipient
  )

```

Histogram of Age Difference
```{r}
liver_data_clean |>
  ggplot(aes(x = age_diff)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Distribution of Donor–Recipient Age Differences",
    x = "Donor Age – Recipient Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```


Calculate the mean and standard deviation
```{r}
age_stats = liver_data_clean |>
  summarise(
    mean = mean(age_diff, na.rm = TRUE),
    sd   = sd(age_diff, na.rm = TRUE)
  )

mu = age_stats$mean
sigma = age_stats$sd

```

Histogram with Normal Distribution Line
```{r}
n <- nrow(liver_data_clean)

liver_data_clean |>
  ggplot(aes(x = age_diff)) +
  geom_histogram(
    binwidth = 2,
    fill = "steelblue",
    color = "white",
    alpha = 0.85
  ) +
  # Add normal curve
  stat_function(
    fun = function(x) n * 2 * dnorm(x, mean = mu, sd = sigma),
    color = "red",
    linewidth = 0.8
  ) +
  # Zero line
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Distribution of Donor–Recipient Age Differences",
    subtitle = paste0("Normal Curve: mean = ", round(mu,1), 
                      ", sd = ", round(sigma,1)),
    x = "Donor Age – Recipient Age (years)",
    y = "Count"
  ) +
  theme_minimal()
```


Donor - Recipient Age Difference based on Living and Deceased Donor 
```{r}
liver_data_clean |>
  filter(don_ty %in% c("C", "L")) |>
  mutate(
    don_ty_label = ifelse(don_ty == "C", "Deceased Donor", "Living Donor")
  ) |>
  ggplot(aes(x = don_ty_label, y = age_diff, fill = don_ty_label)) +
  geom_boxplot(alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Donor–Recipient Age Difference by Donor Type",
    x = "",
    y = "Donor Age – Recipient Age (years)",
    fill = "Donor Type"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

-------------------------------
Exploratory: Does race match?

Make a variable that says if they match
```{r}
liver_data_clean = liver_data_clean |>
  mutate(
    race_match = ifelse(
      don_race_srtr_std == can_race_srtr_std,
      "Matched",
      "Mismatched"
    )
  )
```

Overall Race Match
- Just to be written in the report
```{r}
liver_data_clean |>
  count(race_match) |>
  mutate(pct = n / sum(n))
```

```{r}
liver_data_clean |>
  count(can_race_srtr_std, race_match) |>
  group_by(can_race_srtr_std) |>
  mutate(pct = n / sum(n))
```

Heat Map
```{r}
liver_data_clean |>
  count(can_race_srtr_std, don_race_srtr_std) |>
  group_by(can_race_srtr_std) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(
    x = don_race_srtr_std,
    y = can_race_srtr_std,
    fill = pct
  )) +
  geom_tile(color = "white") +
scale_fill_gradient(
  low = "#eff3ff" ,
  high = "#08519c" ,
  labels = scales::percent
) +
labs(
  title = "Donor–Recipient Race Matching Matrix (%)",
  x = "Donor Race",
  y = "Recipient Race",
  fill = "Percent"
) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


Make variable to see if race matches 
```{r}

race_match_by_don_type <- liver_data_clean |>
  mutate(
    # Clean donor type
    don_ty_label = case_when(
      don_ty == "C" ~ "Deceased Donor",
      don_ty == "L" ~ "Living Donor",
      TRUE ~ NA_character_
    ),
    
    # Define race match variable
    race_match = ifelse(
      don_race_srtr_std == can_race_srtr_std,
      "Matched",
      "Mismatched"
    )
  ) |>
  filter(!is.na(don_ty_label)) |>   # remove any NA donor types
  count(don_ty_label, race_match) |>
  group_by(don_ty_label) |>
  mutate(
    pct = n / sum(n),
    pct_fmt = scales::percent(pct, accuracy = 0.1)
  ) |>
  ungroup()
```


```{r}
race_match_by_don_type_wide <- race_match_by_don_type |>
  select("Donor Type" = don_ty_label, 
         race_match, pct_fmt) |>
  tidyr::pivot_wider(
    names_from = race_match,
    values_from = pct_fmt
  )

race_match_by_don_type_wide |>
  kable(
    format = "html",
    caption = "Race Match Rate by Donor Type",
    align = c("l", "r", "r")
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```







